---
title: "Thema09_EDA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 12,fig.height = 5, warning = FALSE, message = FALSE)
options(warn = 0)
library(ggplot2, quietly = TRUE)
library(gridExtra, quietly = TRUE)
library(kableExtra, quietly = TRUE)
```

# Introduction
Pancreatic ductal adenocarcinoma (PDAC) is one of the deadiest cancers. The chances of survival are increased when diagnosed in an early stage. However, PDAC shows symptoms when it already spread throughout the body. Most of the time, it's too late by then. There may be a way to detect PCAD in an early stage with a simple urine test, with the use of the following biomarkers: creatinine (Urinary biomarker of kidney function) LYVE1 (Urinary levels of Lymphatic vessel endothelial hyaluronan receptor 1, a protein that may play a role in tumor metastasis), REG1A and REG1B (Urinary levels of a protein that may be associated with pancreas regeneration.), and TFF1 (Urinary levels of Trefoil Factor 1, which may be related to regeneration and repair of the urinary tract)

the attributes in the data interesting for this research are the biomarker values mentioned before. There is also an attribute called diagnosis, in which the diagnosis of the sample is stated, where 1 means no PDAC, 2 means benign hepatobiliary disease (non cancerous, non harmful pancreatic condition), and 3 means that the sample has PDAC.

Research question: Is it possible to detect pancreatic cancer using values of the urinary biomarkers?

# EDA
## Codebook
```{r, }
myData <- read.csv("Data/Debernardi et al 2020 data.csv")

columns <- colnames(myData)
type <- c("character", "character", "character", "double", "character", "double", "logical", "logical", "double", "double", "double", "double", "double", "double")
unit <- c(NA, NA, NA, "years", "F/M", NA, NA, NA, "U/ml", "mg/ml", "ng/ml", "ng/ml", "ng/ml", "ng/ml")
descriptions = c("Unique string identifying each subject", "Cohort 1, previously used samples; Cohort 2, newly added samples", "BPTB: Barts Pancreas Tissue Bank, London, UK; ESP: Spanish National Cancer Research Centre, Madrid, Spain; LIV: Liverpool University, UK; UCL: University College London, UK", "Age in years", "M = male, F = female", "1 = control (no pancreatic disease), 2 = benign hepatobiliary disease (119 of which are chronic pancreatitis); 3 = Pancreatic ductal adenocarcinoma, i.e. pancreatic cancer", "For those with pancratic cancer, what stage was it? One of IA, IB, IIA, IIIB, III, IV", "For those with a benign, non-cancerous diagnosis, what was the diagnosis?", "Blood plasma levels of CA 19–9 monoclonal antibody that is often elevated in patients with pancreatic cancer. Only assessed in 350 patients (one goal of the study was to compare various CA 19-9 cutpoints from a blood sample to the model developed using urinary samples).", "Urinary biomarker of kidney function", "Urinary levels of Lymphatic vessel endothelial hyaluronan receptor 1, a protein that may play a role in tumor metastasis", "Urinary levels of a protein that may be associated with pancreas regeneration.", "Urinary levels of Trefoil Factor 1, which may be related to regeneration and repair of the urinary tract", "Urinary levels of a protein that may be associated with pancreas regeneration. Only assessed in 306 patients (one goal of the study was to assess REG1B vs REG1A)")
codebook <- data.frame(columns, type, unit, descriptions)
write.csv(codebook, "Codebook.csv", row.names = FALSE)
knitr::kable(codebook, caption="Table 1: The Codebook", format = 'latex') %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = T) %>%
  column_spec(4, width = "22em")
```
\newpage
## Data exploration
### Visualization
To look at the data and any possible missing data, here's a table containing only the relevant columns for this research.
```{r}
relevant <- myData[c(4, 6, 9:14)]
knitr::kable(head(relevant, n=15), caption="Table 2: Values of biomarkers and the diagnosis") %>%
  kable_styling(position = 'center')
```
\newpage
As is it obvious to see, the REG1A column contains a lot of missing values. The question is exactly how is this possible and what does it mean for the rest of the data? The first and most simple solution is that we don't need this column at all, since the REG1B is similar in function and containing all of the data. Another solution that's not very logical, is to use only the rows where there is a value in the REG1A column. The most logical solution in my opinion is to look at every column separately, and then remove the missing data. When the cleaned data then is plotted, it's possible to look at all the columns together and see if there is any correlation or trend to figure out.


knowing this, let's have a look at the important raw data with the use of boxplots. 
```{r}


p1 <- ggplot(myData, aes(x=age)) +
  geom_boxplot()
p2 <- ggplot(myData, aes(x=creatinine)) +
  geom_boxplot()
p3 <- ggplot(myData, aes(x=LYVE1)) +
  geom_boxplot()
p4 <- ggplot(myData, aes(x=REG1B)) +
  geom_boxplot()
p5 <- ggplot(myData, aes(x=TFF1)) +
  geom_boxplot()
p6 <- ggplot(myData, aes(x=REG1A)) +
  geom_boxplot()

grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 3)
```

as you can see, there are also a lot of outliers which we need to check. If we remove the outliers and clean up the data, without removing the missing values of REG1A, we'll get something like this.

```{r}
Q <- quantile(myData$creatinine, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(myData$creatinine)
up <- Q[2]+1.5*iqr
low <- Q[1]-1.5*iqr

myDataNO <- subset(myData, myData$creatinine > low & myData$creatinine < up)

Q <- quantile(myDataNO$LYVE1, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(myDataNO$LYVE1)
up <- Q[2]+1.5*iqr
low <- Q[1]-1.5*iqr

myDataNO <- subset(myDataNO, myDataNO$LYVE1 > low & myDataNO$LYVE1 < up)

Q <- quantile(myDataNO$REG1B, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(myDataNO$REG1B)
up <- Q[2]+1.5*iqr
low <- Q[1]-1.5*iqr

myDataNO <- subset(myDataNO, myDataNO$REG1B > low & myDataNO$REG1B < up)

Q <- quantile(myDataNO$TFF1, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(myDataNO$TFF1)
up <- Q[2]+1.5*iqr
low <- Q[1]-1.5*iqr

myDataNO <- subset(myDataNO, myDataNO$TFF1 > low & myDataNO$TFF1 < up)

p1 <- ggplot(myDataNO, aes(x=age)) +
  geom_histogram(fill = "lightgrey", col = "black", binwidth = 1)

p2 <- ggplot(myDataNO, aes(x=creatinine)) +
  geom_boxplot()
p3 <- ggplot(myDataNO, aes(x=LYVE1)) +
  geom_boxplot()
p4 <- ggplot(myDataNO, aes(x=REG1B)) +
  geom_boxplot()
p5 <- ggplot(myDataNO, aes(x=TFF1)) +
  geom_boxplot()

grid.arrange(p1, p2, p3, p4, p5, nrow = 3)

```

This looks a lot cleaner, however it is not wise to remove these outliers as they may contain important data in a later stage. 

Before proceeding, it is quite helpful to look at the distribution of the diagnosis column, to see if they're evenly distributed or not.

```{r}
ggplot(myData, aes(x=diagnosis)) +
  geom_histogram(fill = "lightgrey", col = "black", binwidth = 1) +
  ggtitle("Distribution of the diagnosis")

```
Now we can check if the different biomarkers correlate with each other in any way.

```{r}
p1 <- ggplot(myData, aes(x=creatinine, y=LYVE1)) +
  geom_point() +
  ggtitle("Creatinine vs LYVE1")
p2 <- ggplot(myData, aes(x=creatinine, y=REG1B)) +
  geom_point() +
  ggtitle("Creatinine vs REG1B")
p3 <- ggplot(myData, aes(x=creatinine, y=TFF1)) +
  geom_point() +
  ggtitle("Creatinine vs TFF1")
p4 <- ggplot(myData, aes(x=LYVE1, y=REG1B)) +
  geom_point() +
  ggtitle("LYVE1 vs REG1B")
p5 <- ggplot(myData, aes(x=LYVE1, y=TFF1)) +
  geom_point() +
  ggtitle("LYVE1 vs TFF1")
p6 <- ggplot(myData, aes(x=TFF1, y=REG1B)) +
  geom_point() +
  ggtitle("TFF1 vs REG1B")

grid.arrange(p1, p2, p3, p4, p5, p6, nrow=3)
```
I think it's safe to assume that none of these correlate with each other in any way. One thing to notice however, is that LYVE1, REG1B and TFF1 have a lot of values close to zero. We can figure out if this is important or not in a future stage of the research. 

To ensure that all the values are completely independant we can perform a principal component analysis.
```{r}

```


What we should do now, is look at the values of those urinary levels with each diagnosis and see if there are any obvious differences.  
Let's start with the samples that do not have PDAC
```{r}
myData1 <- subset(myData, myData$diagnosis == 1)

p1 <- ggplot(myData1, aes(x=age)) +
  geom_histogram(fill = "lightgrey", col = "black", binwidth = 1) +
  ggtitle("Age distribution")

p2 <- ggplot(myData1, aes(x=creatinine)) +
  geom_boxplot() +
  ggtitle("Distribution of creatinine")
p3 <- ggplot(myData1, aes(x=LYVE1)) +
  geom_boxplot() +
  ggtitle("Distribution of LYVE1")
p4 <- ggplot(myData1, aes(x=REG1B)) +
  geom_boxplot() +
  ggtitle("Distribution of REG1B")
p5 <- ggplot(myData1, aes(x=TFF1)) +
  geom_boxplot() +
  ggtitle("Distribution of TFF1")
p6 <- ggplot(myData1, aes(x=REG1A)) +
  geom_boxplot() +
  ggtitle("Distribution of REG1A")

grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 3)
```

Here are the values of the samples with non-cancerous pancreatic conditions.

```{r}
myData2 <- subset(myData, myData$diagnosis == 2)

p1 <- ggplot(myData2, aes(x=age)) +
  geom_histogram(fill = "lightgrey", col = "black", binwidth = 1) +
  ggtitle("Age distribution")

p2 <- ggplot(myData2, aes(x=creatinine)) +
  geom_boxplot() +
  ggtitle("Distribution of creatinine")
p3 <- ggplot(myData2, aes(x=LYVE1)) +
  geom_boxplot() +
  ggtitle("Distribution of LYVE1")
p4 <- ggplot(myData2, aes(x=REG1B)) +
  geom_boxplot() +
  ggtitle("Distribution of REG1B")
p5 <- ggplot(myData2, aes(x=TFF1)) +
  geom_boxplot() +
  ggtitle("Distribution of TFF1")
p6 <- ggplot(myData2, aes(x=REG1A)) +
  geom_boxplot() +
  ggtitle("Distribution of REG1A")

grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 3)
```

And finally the samples that do have PDAC.

```{r}
myData3 <- subset(myData, myData$diagnosis == 3)

p1 <- ggplot(myData3, aes(x=age)) +
  geom_histogram(fill = "lightgrey", col = "black", binwidth = 1) +
  ggtitle("Age distribution")

p2 <- ggplot(myData3, aes(x=creatinine)) +
  geom_boxplot() +
  ggtitle("Distribution of creatinine")
p3 <- ggplot(myData3, aes(x=LYVE1)) +
  geom_boxplot() +
  ggtitle("Distribution of LYVE1")
p4 <- ggplot(myData3, aes(x=REG1B)) +
  geom_boxplot() +
  ggtitle("Distribution of REG1B")
p5 <- ggplot(myData3, aes(x=TFF1)) +
  geom_boxplot() +
  ggtitle("Distribution of TFF1")
p6 <- ggplot(myData3, aes(x=REG1A)) +
  geom_boxplot() +
  ggtitle("Distribution of REG1A")

grid.arrange(p1, p2, p3, p4, p5, p6, nrow = 3)
```

At first glance, there are notable differences but we can further investigate this in future steps of this research.

```{r}
write.csv(myData, "Data/DataCleaned.csv")
```

